--- 
title: "Introduction to R"
author: Gaby Mahrholz, Greta Todorova, Wilhelmiina Toivo, and Jaimie Torrance, updated by Johanna Werner in 2022 and Sean Westwood in 2023
date: "`r Sys.Date()`"
bibliography: [book.bib, packages.bib]
biblio-style: "apalike"
description: This is the course handbook for the Glasgow University Open Studies course Introduction to R, academic session 2023/2024.
documentclass: book
link-citations: yes
site: bookdown::bookdown_site

---

# About this course {-}

```{r setup, include = F}
# to render entire book run: bookdown::render_book(input = input_files) in the console with this object in the environment
input_files <- list.files()[endsWith(list.files(), ".Rmd")] 

knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse'
), 'packages.bib')
```

```{r setup2, include=FALSE}
library(tidyverse)
library(knitr)
library(webexercises)

add_to_bookdown()

# knitr::opts_chunk$set(echo = TRUE)
# if (!requireNamespace("webexercise")) {
#   stop("You must have the 'webex' package installed to knit HTML from this template.\n   devtools::install_github(\"psyteachr/webexercise\")")
# } else {
#   library("webexercise")
# }

```


This course is a short introduction on using R for dealing with data. 

## Intended Learning Outcomes {-}

The Intended Learning Outcomes (ILOs) for the entire course are:

* Demonstrate how R and RStudio work by transforming and cleaning different types of data into an appropriate format for analysis and creating graphical representations to visualise different types of data;
* Calculate summary statistics using R; and
* Interpret common error messages and use in-built help facilities to debug their code. 

Each chapter contained within will come with its own specific ILOs which will be clearly specified at the top of each page. 

These ILOs will be the points that will be covered in the assessments.

## What will I need to complete this course? {-}

To successfully complete this course, you will need access to a computer that either already has, or you can download R and RStudio software on. This computer will need to use a Windows, Mac or Linux based operating system. You *can* run RStudio on a Chromebook sometimes, but the process is more complex (see <a href="https://www.reddit.com/r/chromeos/comments/8r5lwb/r_studio_for_chromebook/">here</a>).

If you cannot use RStudio on your computer you might want to consider using <a href="https://rstudio.cloud/">RStudio Cloud</a>.

All University of Glasgow Library computers should have this software installed. If you have enrolled in the University, you should have access to these.

Before the first lesson you will be given the task of either installing R and RStudio on your own personal computer, or finding another suitable method of accessing the software. Full instructions will be given in [Chapter 1](\@ref(#accessing)) of this coursebook.

```{block, type = "info"}
**If you are having issues with access, talk to your tutors via the via the Microsoft Teams channel as soon as possible so that we can help you!** R, like any skill, will get better with practice and we want to ensure you can practice and complete the assessments as easily as possible.
```



## Why are these materials open? {-}

These materials have been adapted from various sources. Some of these materials have been Open Educational Materials with CC-BY-SA licenses. This means that any materials that use them must also be made open (SA means "share alike"). 

We also believe in the value of open educational materials. Your course fees contribute to the development of this course, but also to providing you with in-person classes and supporting staff.

## Acknowledgement {-}

The materials were initially developed for the University of Glasgow Short Courses for the academic term 2019/20 by [Gaby Mahrholz](https://www.psy.gla.ac.uk/otherstaff/?id=GM005), [Kate Haining](https://www.psy.gla.ac.uk/otherstaff/?id=KH001), [Rebecca Lai](https://www.psy.gla.ac.uk/otherstaff/?id=RL003), [Greta Todorova](https://www.psy.gla.ac.uk/otherstaff/?id=GKT01), and [Wilhelmiina Toivo](https://www.psy.gla.ac.uk/otherstaff/?id=TIWT1), and modified for the academic term 2023/24 by [Sean Westwood](https://www.gla.ac.uk/schools/psychologyneuroscience/staff/seanwestwood/)
