---
title: "L6 inclass"
author: "SW"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tidy Data 2: Cell Manipulation 

## Pre-steps

Before we can start to focus on some other functions that might come in handy for data manipulation, we need to make sure to load `tidyverse` into the library.

```{r L6}
library(tidyverse)
```



## Conditional statements

Conditional statements are used when you want to create an output value that is conditioned on an evaluation. Here, we are showing you how to use them in combination with `mutate()` by either modifying the values within one column, or by adding a new column.

Example scenario: Imagine that we have data in which a column `Sex` holds values of `1` and `2` that stand for `Male`, and `Female` participants respectively. Conditional statements make an evaluation: *if `Sex` is `1`, then put the value of `Male` in a new column called `Sex_char`; if `Sex` is `2` put `Female` in column `Sex_char`.*

Let's create some fake data for this. 40 participants (17 male, 20 female, 3 other), they are between 16 and 40 years of age, they study at 4 different institutions (University of Glasgow, University of Strathclyde, Glasgow Caledonian University, University of the West of Scotland), and we had them complete an R assignment and recorded their grade points.

```{r fake_data}
set.seed(999) #so we all get the same random data
fake_data <- tibble(PP_ID = 1:40,
                    Sex = c(rep(1,17), rep(2,20), rep(3,3)),
                    Age = sample(16:40, 40, replace = TRUE),
                    Institution = sample(c(rep(1,10), rep(2,8), rep(3,10), rep(4,10), rep(99,2))),
                    Grade_points = sample(c(0,9:22), 40, replace = TRUE))
head(fake_data)

```


We can use the functions `recode()`, an `ifelse` statement or `case_when()` to recode values to make the data more meaningful.


### recode()

Can be used for character, numeric, and factors columns.

The structure for `recode()` is:

```{r, eval=FALSE}
recode(variable_to_recode, old_value = new_value)
```

This is a bit tricky to remember as it doesn't follow the usual tidyverse logic. In `select()`, `rename()`, and `mutate()` for example it is the other way around `new_thing = old_thing` or `new_thing = something_computational`)

In combination with `mutate()`, the structure looks as follows:

```{r, eval=FALSE}
mutate(data, new_column_name = recode(variable_to_recode, old_value = new_value))
```

We need a new column name within the `mutate()` function but also a reference column within `recode()` on which the conditional recoding is based upon.

So, let's turn to our `fake_data`, and apply that logic for recoding the values of the column `Sex` where 1 = "Male", 2 = "Female", and 3 = "Other". We will add a new column `Sex_char` onto our `fake_data`.

Recoding numeric values requires backticks (if the numbers are on the left side of the equation).

```{r}
fake_data <- mutate(fake_data, Sex_char = recode(Sex, 
                                                 `1` = "Male", 
                                                 `2` = "Female", 
                                                 `3` = "Other"))
head(fake_data)
```



**Your turn**

The next variable to recode is `Institution`:

* 1 = UoG, 
* 2 = Strathclyde, 
* 3 = GCU, 
* 4 = UWS

In the questionnaire, two participants forgot to indicate which university they are studying at (coded as 99), but fortunately we know that they are at the University of Strathclyde (the test was distributed to 10 students at each institution, however, Strathclyde only has 8 responses).

Add a new column `Institution_char` to `fake_data` that holds the recoded values as the character strings indicated above.


```{r}
fake_data <- NULL
```



An alternative to `recode()` is `ifelse()`. Let's see how `Sex` and `Institution` could have been recoded using `ifelse()`.



### ifelse()

`ifelse()` is a good option when dealing with character or numeric input but struggles with dates and factors. None of the columns in `fake_data` are dates or factors, so we can easily use `ifelse()` for recoding variables.

The generic structure for `ifelse()` is:

```{r, eval=FALSE}
ifelse(test, yes, no)
```

* test: is an expression to test that creates an output of `TRUE` or `FALSE`
* yes: if test has a `TRUE` outcome, this is what would be coded in the new column
* no: if test has a `FALSE` outcome, this is what would be coded in the new column

The interpretation of `ifelse()` is: if the test expression is `TRUE`, write the value of yes, otherwise no.

In combination with `mutate()`, the structure looks as follows:

```{r, eval=FALSE}
mutate(data, new_column_name = ifelse(test, yes, no))
```

How would we apply that to recoding `Sex`? We would want the expression to test whether the Sex column holds a value of 1, then a new column should code it as "Male", otherwise as "Female". Let's add a new column `Sex_ifelse` to `fake_data`

```{r}
fake_data <- mutate(fake_data, Sex_ifelse = ifelse(Sex == 1, "Male", "Female"))
head(fake_data)
```


But now the 3 participants that chose option 3 are coded as "Female". Solution: a nested `ifelse()` statement. Instead of replacing the no argument (`FALSE` evaluation) with "Female", we could include a second `ifelse()` statement.

```{r}
fake_data <- mutate(fake_data, Sex_ifelse = ifelse(Sex == 1, "Male",
                                                   ifelse(Sex == 2, "Female", "Other")))
```

Now everything is coded as it should be.


**Your turn**

Recode `Institution` using `ifelse()`. Add a new column `Institution_ifelse` to `fake_data`. A quick reminder of the `Institution` values:

* 1 = "UoG", 
* 2 and 99 = "Strathclyde", 
* 3 = "GCU", and
* 4 = "UWS"


```{r}
fake_data <- NULL
```
`r unhide()`


The test expression in `ifelse()` can evaluate more complex "things" as long as they return a logical value. For example, if we wanted to add a column `Pass_Fail` to `fake_data`, that holds values of "Pass" for grade points above 9, and "Fail" for values of 9 and below, we would code


```{r}
fake_data <- mutate(fake_data, Pass_Fail = ifelse(Grade_points > 9, "Pass", "Fail"))
head(fake_data)
```

We could also add a column `Grade` to `fake_data` that tells us the primary band the students got in their test. `Grade_points` between 18 and 22 result in A, 15 to 17 = B, 12 to 14 = C, 9 to 11 = D, 6 to 8 = E, 3 to 5 = F, 1 and 2 = G, and 0 = H. Here we can use the `%in%` operator.

```{r}
fake_data <- mutate(fake_data, Grade = ifelse(Grade_points %in% 18:22, "A",
                                              ifelse(Grade_points %in% 15:17, "B",
                                                     ifelse(Grade_points %in% 12:14, "C",
                                                            ifelse(Grade_points %in% 9:11, "D",
                                                                   ifelse(Grade_points %in% 6:8, "E",
                                                                          ifelse(Grade_points %in% 3:5, "F",
                                                                                 ifelse(Grade_points %in% 1:2, "G","H"))))))))

head(fake_data)
```


### case_when()

`case_when()` vectorises multiple `ifelse()` statements, and might potentially be a touch easier to digest visually than `ifelse()`.

There are few guidelines to follow when using `case_when()`:

* A sequence of two-sided formulas is needed. 
* The left hand side (LHS) determines which values match this case. The right hand side (RHS) provides the replacement value.
* The LHS must evaluate to a logical vector. The RHS does not need to be logical, but all RHSs must evaluate to the same type of vector.

```{r, eval=FALSE}
mutate(data, new_column_name = case_when(test ~ new_value))
```

Let's apply this to recoding the values in the column `Institution`. Combining `case_when()` with `mutate()`, we will add another column called `Institution_case_when` to `fake_data`.

```{r}
fake_data <- mutate(fake_data, Institution_case_when = case_when(
  Institution == 1 ~ "UoG",
  Institution == 3 ~ "GCU",
  Institution == 4 ~ "UWS",
  TRUE ~ "Strathclyde"
  )
)
head(fake_data)
```

The `TRUE` statement at the end is similar to the the "else" in the `ifelse()` function. A further similarity between the two is that arguments are evaluated in order, so you must proceed from the most specific to the most general.



**Your turn**

Add a column `Grade_case_when` into `fake_data` using a combination of `mutate()` and `case_when()`, that recodes grade points to primary bands. Quick reminder:

* grade points 18 - 22 = A,
* 15 to 17 = B, 
* 12 to 14 = C, 
* 9 to 11 = D, 
* 6 to 8 = E, 
* 3 to 5 = F, 
* 1 and 2 = G, and 
* 0 = H.




```{r}
fake_data <- NULL
```



## distinct()

The `distinct()` function is used to remove duplicate rows in your dataframe. If we are using `distinct()` without specifying any arguments, R checks the whole row and excludes any row that is exactly repeated.

```{r}
fake_data_no_dup <- fake_data %>% 
  distinct()
```

No duplicates were removed here which is not surprising, given every participant has a unique ID number. However, we can specify variables we want to see unique values for. Say we we wanted to see how many different institutions we have data from.


```{r}
institution_distinct <- fake_data %>% 
  distinct(Institution_char)
institution_distinct
```

Additional argument: `.keep_all`. If we set that to `TRUE` (the default here is `FALSE`), we can keep all the other columns in the dataset.


```{r}
institution_distinct <- fake_data %>% 
  distinct(Institution_char, .keep_all = TRUE)
institution_distinct
```

`distinct()` only preserves the first row a new value occurs.

The `distinct()` function also allows you to eliminate duplicate rows with multiple variables in the dataframe. So if we wanted to see where participants were studying but also the different sexes taking part in the study, we would code...

```{r}
institution_sex_distinct <- fake_data %>% 
  distinct(Institution_char, Sex_char)
institution_sex_distinct
```

`institution_sex_distinct` has 10 observations, so that means, we have at least 1 male and 1 female participant from each of the 4 institutions. Additionally, Strathclyde and Glasgow have participants classifying as neither male nor female (here coded as "Other"). 


**Question**
You can see they are not sorted by either of the column. Why is that?

Answer:



## Recap

Let's just have some fun with `fake_data` whilst recapping a little bit of what we did in the past 5 weeks.

### Question 1

Select the variables `PP_ID`, `Sex_char`, `Age`, `Institution_char`, `Grade_points`, `Grade`, and `Pass_Fail` from `fake_data`and save it as `fake_data` in your Global Environment (i.e. overwrite/replace `fake_data`). Rename the following variables in the process:

* `PP_ID` as `Participant_ID`
* `Sex_char` as `Sex`
* `Institution_char` as `Institution`

```{r}
fake_data <- NULL
```



### Question 2

Calculate the average grade points for each University. Store the output as a new object `grade_points` in your `Global Environment`. The new column should be named `Average_grade_points`.

```{r}
grade_points <- NULL
```




### Question 3

Join the `fake_data` and `grade_points`, and calculate how much each person differs from the mean of their university. Results should be stored in a column named `grade_diff`. Store the output in your `Global Environment` as a new object called `fake_data_meandiff`. Sort the data by `Institution` and `Sex`.

```{r}
fake_data_meandiff <- NULL
```




### Question 4

Our colleague is interested in what the grade point average for each university would be if we were only looking at the students who passed the R assignment initially (i.e. grade points above 9). Store the output in your `Global Environment` as a new object called `fake_data_pass`.

```{r}
fake_data_pass <- NULL
```



### Question 5

Now our colleague wants to know whether the males or the females did better at each university. They are asking for a table with 4 columns in which the first column titled `Institution` lists the university, the second column is called `Males` for the average grade point scores of the males at each institution, the third column called `Females` with average grade point scores for the females, and the fourth column `Better` should say `"female"` or `"male"` if the females or males had higher grade points respectively. Store the output in your `Global Environment` as a new object called `fake_data_sex`. Use `fake_data` as the reference dataframe.

```{r}
fake_data_sex <- NULL
```




# Tidy Data 3: Joins 

## Pre-steps

Before we can start we need to make sure to load `tidyverse` into the library. We're also going to use the `babynames` data again so let's load that.

```{r L6, message=FALSE}
library(babynames)
```

## Two-Table Verbs

More often than not you will find yourself with data across multiple tables. In order to effectively combine multiple tables, `dplyr` provides a selection of **mutating join verbs**.

### Bringing in new data

First of all lets bring in some familiar data to look at these new functions.

#### What's the data

We're going to use 2 datasets that you should be familiar with from last week. The `starwars` data from the `tidyverse` and the `babynames` data package.

```{r}
starwars
```

```{r}
babynames
```

#### Mutating Join Verbs

Mutating join verbs combine the variables (columns) of two tables so that matching rows are together. There are 4 different types of mutating joins, namely `inner_join()`, `left_join()`, `right_join()`, and `full_join()`.

I'm going to adapt the starwars data to make it a bit easier for us to use.

```{r sw_data}
sw_names <- starwars %>%
  separate(col = name, into = c("name", NA), sep = " ", extra = "drop", fill = "right") %>%
  unnest(films) %>% # Don't worry, we didn't learn about this
  filter(films == "The Empire Strikes Back") %>%
  mutate(sex = recode(sex, male = "M", female = "F", .default = NA_character_)) %>%
  select(name, sex, homeworld) %>%
  arrange(name)

sw_names
```
Now we have all the first names of the characters in The Empire Strikes Back. Notice that some of these names will be in the babynames dataset and some won't.

##### inner_join()

`inner_join()` returns all rows from both tables for which the values in column specified in the `"by = "` statement overlap.

Here we merge `babynames` and `sw_names` with `inner_join()` specifying we want to "join by" the variable `name`. Because both tables have the column `sex`, R needs to distiguish them from each other, for that we use the `suffix = ` argument, which allows us to specify what suffix to add to each column of the resulting table. If we do not spcify this, R will add a `.x` and `.y` suffix for us to the x (left) and y (right) table variables respectively.

```{r}
inner <- inner_join(sw_names, babynames, by = join_by(name), suffix = c("_sw", "_baby"))
inner
```

Using an inner join returns only `r nrow(inner)` rows of observations because these are the only rows that contain names from our Star Wars data. However, we are still merging both tibbles together, meaning that all columns from `sw_names` and `babynames` are kept (including the `sex` associated with that name in each table, with a suffix added to each to tell them appart). In this example, the name "Yoda" is dropped because no one is actually called "Yoda".

```{r}
filter(inner, name == "Yoda")
```

##### left_join()

`left_join()` retains the complete first (left) table and adds values from the second (right) table that have matching values in the column specified in the `"by ="` statement. Rows in the left table with no match in the right table will have missing values (`NA`) in the new columns. 

Let's try this `left_join()` function for our simple example of `babynames` and `sw_names` in R.

```{r}
left <- left_join(sw_names, babynames, by = join_by(name), suffix = c("_sw", "_baby"))
left
```

Here `babynames` is returned in full, and for every matching `name` in `sw_names` the value is added. However, most of the names in `babynames` aren't in `sw_names`, so we get `NA` added here to the column `sex_sw` and `homeworld` columns.  

We can get the same data as we did with `inner_join()` if we remove the missing values. We can check if 2 objects are the same in R using the `identical()` function.

```{r}
left_filt <- filter(left, !is.na(year))

identical(left_filt, inner)
```

If we want to do it the other way round we can reverse the order of the tables. This time I'll also leave out the `by` argument so that we only match the rows where both `name` and `sex` match. (Note the helpful message we get.)

```{r}
left_join(babynames, sw_names)
```

##### right_join()

Alternatively we can use a `right_join()` to do the same thing using the original order of data tables. `right_join()` retains the complete second (right) table and adds values from the first (left) table that have matching values in the column specified in the by statement. Rows in the right table with no match in the left table will have missing values (`NA`) in the new columns.

```{r}
right <- right_join(sw_names, babynames, by = join_by(name, sex))
right
```

Here `sw_names` is returned in full, and for every matching `name` in `babynames` the value is added, for any row of `sw_names` that does not have a matching value in `babynames`, `NA` is added. Notice the order of the rows, though!!! All the names from `sw_names` come first before the extra rows from `babynames` which are added at the bottom. That is due to the order of how they are entered into the `right_join()` function. The "left" data (first table) is still prioritised in terms of ordering observations!

```{r}
arrange(right, name)
```

##### full_join() #####

`full_join()` combines all rows and all columns from both dataframes. `NA` values fill unmatched rows.

```{r}
full <- full_join(sw_names, babynames, join_by(name), suffix = c("_baby", "_sw"))
full
```

As you can see, all names from both tables are kept, and `NA` is used to fill the missing `homeworlds`. Again you can see the prioritization of the left (first) table in the arrangement of the years.

### Mutating Join Summary

| join function | Description                                                             |
|:--------------|:------------------------------------------------------------------------|
| inner_join()  | Includes all rows that are PRESENT IN BOTH the left and the right table |
| left_join()   | Includes all rows from the left table (first data entered)              |
| right_join()  | Includes all rows from the right table (second data entered)            |
| full_join()   | Includes all rows from both left and right tables                       |

Let's make a new tibble for you to use. Make sure you run the code in this block.

```{r}
beatnames <- tibble(
  name = c("John", "Paul", "George", "Ringo", "Yoko", "Walrus"),
  n = 1:6
)
```

Join together `beatnames` and `babynames` so that we keep all the rows from the `beatnames` table, add a useful suffix so you can differentiate between columns with the same name. Try using each of the 4 mutating join methods and store the result in `beat_left`, `beat_right`, `beat_inner` and `beat_full`.

```{r}
beat_left <- NULL
```

```{r}
beat_right <- NULL
```

```{r}
beat_inner <- NULL
```

```{r}
beat_full <- NULL
```

Can you do something to get the same rows as `beat_inner` from `beat_full`?

```{r}
beat_inner2 <- NULL
```

###	Binding Join Verbs

In contrast to mutating join verbs, binding join verbs simply combine tables without any need for matching. Dplyr provides `bind_rows()` and `bind_cols()` for combining tables by row and column respectively. When row binding, missing columns are replaced by `NA` values. When column binding, if the tables do not match by appropriate dimensions, an error will result. 


#### bind_rows()

```{r}
wos_seasonal_sun <- read_csv("data/wos_seasonal_sun.csv", show_col_types = FALSE)
wos_monthly_sun <- read_csv("data/wos_monthly_sun.csv", show_col_types = FALSE)
```


`bind_rows()` is ideal when you have more entries of the same kind of data, i.e. new observations of the same variables. (For example; you have a new batch of participants answering the same questionnaire; or you have new air pollution data from a different geographic region... same variables - different observations)

Lets split some data

```{r}
Sun_Season_1 <- wos_seasonal_sun %>%
  filter(year >= 1970)

Sun_Season_2 <- wos_seasonal_sun %>%
  filter(year < 1970)
```

Now we have 2 identical tables for seasonal sunshine in the west of Scotland, but for different sets of years. All the same variables, but totally different observations.

We can easily join these together with `bind_rows` to create a complete history.
```{r}
Bind_1 <- bind_rows(Sun_Season_1, Sun_Season_2)
Bind_1
```
`bind_rows()` takes the second table `Sun_Season_2` and puts it *directly underneath* the first table `Sun_Season_1`. 


What happens if we attempt to bind tables with different dimensions
```{r}
Bind_2 <- bind_rows(Sun_Season_1, inner)
Bind_2
```
Notice that the `bind_rows()` does not "care" if it duplicates rows, here we have repeat years. Also `bind_rows()` does not "care" that there are columns that do not match between the tables, here `NA` is added fill the missing space. 

#### bind_cols()

`bind_cols()` is similar to our mutating join verbs in that it creates a new table by joining columns (or variables) together. However, note that `bind_cols()` does not perform any kind of row-wise matching when joining tables.

```{r}
Bind_3 <- bind_cols(wos_monthly_sun, wos_seasonal_sun)
```
`bind_cols()`takes `wos_seasonal_sun` and puts it *right next* to `wos_monthly_sun`. Since the column name `year` is in both tables they get called `year..1` and `year...15` after their respective column number to differentiate them.



What happens if we attempt to bind tables with different dimensions
```{r, error=TRUE}
Bind_4 <-bind_cols(wos_monthly_sun, inner)
```
You simply get an error

Just to further clarify the differences between `bind_cols()` and the mutating joing verbs, lets look at how they would deal with the same challenge.

First lets create a new table for winter sun data, but lets arrange it by "win", so that rather than being in order of `year`, it is in order of most sunshine. This will have the same dimensions as the Summer_Sun table but a totally different order

```{r}
Winter_Sun <- wos_seasonal_sun %>%
  select(year, win) %>%
  arrange(desc(win))
Winter_Sun
```

Now lets make a `Summer_sun` table the same way and try joining it with `Winter_Sun` using the two different methods.

```{r}
Summer_Sun <- wos_seasonal_sun %>%
  select(year, sum) %>%
  arrange(desc(sum))


Comparison_1 <- bind_cols(Summer_Sun, Winter_Sun)
Comparison_1

Comparison_2 <- left_join(Summer_Sun, Winter_Sun, by = join_by(year))
Comparison_2
```
