---
title: "Visualisations"
author: "Greta Todorova & Sean Westwood"
date: "03/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Data Visualisation 

Firstly, load the the `tidyverse` package. 

```{r warning=FALSE, echo=FALSE}
library(tidyverse)
```

Today, we will continue working with ggplot2 which is part of tidyverse. We will continue working with the data from the Scottish Government saved in the file free_movement_uk.csv.


## Additional Resources


* [Cookbook for R - Graphs](http://www.cookbook-r.com/Graphs/)
* Hadley Wickham et al.'s [Elegant graphics for Data Analysis](https://ggplot2-book.org/index.html)
* [ggplot2 cheatsheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
* [ggplot2 functions reference](https://ggplot2.tidyverse.org/reference/#section-plot-basics)

## Intended Learning Outcomes

1. Develop an understanding of the basics of ggplot2 and how layering works
2. Learn how to make simple two layer plots
3. Learn how to make plots with more than 2 layers
4. Learn how to create groups in plots
5. Learn how to make a variety of plots based on the type of variables we have
6. Learn the difference between global and local aesthetics
7. Learn how to make multiple plots at the same time
8. Learn how to work with themes: adding labels, changing font size, colours and axes ticks.
9. Learn to work with legends
 
## Pre-steps

We will be working with `ggplot2` today which is part of `tidyverse`.

Firstly, load the `tidyverse` package.  
```{r, message=FALSE}
library(tidyverse)
```


Next, we will be working with data from the Scottish Government saved in the file `free_movement_uk.csv`. Read the data into your `Global Environment` as `migration_scot`.


```{r}
migration_scot <- read_csv('free_movement_uk.csv') %>%
  rename(Migration_Type = `Migration Type`,
         Migration_Source = `Migration Source`)

```

##Introduction to the data

This data is openly available from the Scottish Government, and it introduces the flow of people at different ages and sex into and out of the Scotland from the rest of UK (RUK) and Overseas. We have several variables to work with:

FeatureCode =  Codes given by the Scottish Government
  DateCode = Year of data collected
  Measurement = What type of measurement it is (here we have only counts i.e. the number of people)
  Units = Units (here we have number of people)
  Value = The actual count (i.e. the number of people)
  Age = Age of the counted people (separate by age, and total (sum of all age groups))
  Sex = Sex of the counted people (separate by sex, and total (sum all sex groups))
  `Migration Source` = Where the people are coming from (Overseas, RUK)
  `Migration Type` = Whether people are coming or leaving (In, Out, and Net (people coming in less the people leaving))



The data is a bit messy so lets clean it a little bit before we can work with it. 

### Some warm-up tasks 

Use pipes `%>%` for the following 4 tasks and save your data into the **traffic_scot** variable so that it appears in your `Global Environment`.

1. Select only the necessary variable - `DateCode`, `Value`, `Sex`, `Age`, `Migration Source` & `Migration Type` (notice that some variable names include spaces)
2. Filter the data in column `DateCode`, and keep only observations from 2016
3. Keep only the total number of people by `Age` ('All') and `Sex` ('Male' & 'Female')
4. Keep only the information about the number of people that came into and left Scotland (we do not want to know about the net value)


```{r}
traffic_scot <- NULL
```

## The package GGPLOT2

We use the `ggplot2` package due to its ease of layering and variety. With `ggplot2` we can create beautiful and varied plots that can show us much more than just the simplicity of the base R plots. Whereas the plot/graph's job is to visualise the data, how you visualise your data is very important. Sometimes in an attempt to show off our plotting skills, we can make the graph difficult to read. Today and next week, using `ggplot2` we will streamline the process of plot creation. 


## Layering

Plots created with ggplot work by using layers. 

In the simplest form we have two layers. 

 Layer1: what data are we going to use
      +
 Layer2: how do we want it to look

Layer 1: ggplot(data, aes(x,y, ...)) + 
Layer 2:   geom_something(...)



### Layer 1

Plots always start with **ggplot()**. This initialises the plot. It is used to tell **R** what data we are going to use. We can also set the aesthetics of the plot which will be used to specify the axes of our plot, any groupings, variables we are going to colour based on, etc. Unless specifically overwritten in the following layers, all the data information you give in this first layer will be inherited in every following layer.

For now we are going to stick with one global specification - i.e. we are going to specify all our data in this first layer.

If you run just the first command, R is going to show you the first layer - i.e. a blank screen. R has prepared a space for you to create your plot.

Let's try it.

Run the command bellow and see what appears in the 'Plots' section of R.

```{r}
ggplot()

```

Now let's add some information. Let's include the data and the axes. From our data `traffic_scot` let's put `Sex` on the x-axis and `Value` on the y-axis. This way we can visualise how many people from each sex category came to and left Scotland overall.

What's the difference?

```{r}
ggplot(traffic_scot, aes(x = Sex, y = Value))

```

**Question: What is the difference between the above two lines? What does the first layer give you?**


### Layer 2 and onwards

From the second layer onward, we are specifying how we want our plot to look: from the type of graph to the titles and colours. For now, we are going to stick with the basics. 



## How do we choose the type of plot?

There are numerous plots we can choose, and we are going to try several today. 

Before plotting we need to know what types of data we want to portray. 

Depending on whether our data is discrete or continuous will change what type of plot is most appropriate.


### Data format

#### Continuous

Continuous data has no bounds and the distance between any two consecutive points is equivalent to the distance between any other two consecutive points - i.e. the difference between 1 and 2 is the same as the difference between 3 and 4, or 5 and 6 etc. If you think back to `L2`, integer and double data types will fall in this category.  

#### Categorical

Categorical data, simply put, groups (categorises) data. The data in each group or category vary solely on one characteristic and usually those characteristics cannot be ranked - i.e. sex, race, nationality etc.  If you think back to `L2`, logical and character data types will fall in this category.


### Number of variables

Next we need to know the relationship between how many variables we want to show. In a 2D plot, we can usually graphically show between 1 and 3 variables with ease. After that, **usually** more than 2 dimensions are necessary. 

Let's get back to our free movement data.

Q: How many variables have we chosen so far?
**Write your answer here**
A: 


Q: What types of data are we trying to plot according to our first layer? Continuous, categorical, or both?
**Write your answer here**
A: 


### Types of plots

There is variety of plots that we can create using ggplot2. We select the plot by adding on `geom_???()`. Instead of the **???** we specify the type of plot we want to make. Bellow are some examples, but you can see more here: ggplot2 layers.

| Geom | Plot type  |
|:-----------------|:---------------------------------------------------------|
| `geom_col()` `geom_bar()` | Bar plots  |
| `geom_boxplot()` | Box with whiskers plot |
| `geom_density()` | Density plot |
|	`geom_histogram()` | Histogram |
|	`geom_point()` | Scatter plot|
|	`geom_jitter()` | Jittered points plot |
| `geom_line()` | Line plot |
| `geom_violin()` | Violin plot |
| `geom_smooth()` | Smoothed conditional means (often used to add a regression line) |


Create a column/bar plot to show the data use `geom_col()`. How does the data look like?

```{r}
ggplot(traffic_scot, aes(Sex, Value)) + 
  geom_col()
```


Q: Considering what we know from the data description and the different variable in it, do we need another variable to make the plot more informative? If yes, which one?

**Write your answer here**
A: 


### Groupings

When we want to group data to provide a better representation, we can usually use colour, shapes etc. Most of the time, this will depend on the type of graph we want to create. If we are using line graphs, then colours are usually best. If we are using dot plots then we can use the shape as well as the colours to differentiate the different groups.

Let's first look at colour.

Depending on the type of graph, colouring in can be done either with `colour` or with `fill`. For our column plots we use `fill.` If we used `colour`, then we would get different colours for the outlines. You can find out whether to use colour or fill by typing the name of the geom in the Help search bar or typing `?geom_col()` in the console and looking at the **Aesthetics** section.

```{r}
ggplot(traffic_scot, aes(x = Sex, y = Value, fill = `Migration Type`)) + 
  geom_col(position = 'dodge')
```
Let's decompose the code above: 

We have our data and our two variables. They are the same as before. This time we are saying that we want the data to be separated even further by `Migration Type` and we have given two different colours to the 'In' and 'Out' migration types using the additional argument `fill`. 

We have also added the argument `position = 'dodge'` in the `geom_col()`. This makes sure that any groupings we create are then presented separately.

Simple task: 

1. Remove `position ='dodge'` from the geom and describe what happens. 


```{r}
#create your plot here
ggplot(traffic_scot, aes(Sex, Value, fill=`Migration Type`)) +
  geom_col()
```


We have also added `position = 'dodge'` in the `geom_col()`. This makes sure that any groupings we create and create new columns are then presented separately.

Simple task: 
2. Use `position = 'fill'` and describe what happens. 


```{r}
#create your plot here
ggplot(traffic_scot, aes(Sex, Value, fill=`Migration Type`)) +
  geom_col(position='fill')
```
**Write your explanation of the differences between 'dodge', 'fill' and no position argument here. (Use `?geom_col()` to help you)**

The argument `position` is an argument specific for the `geom_col`. It is just one example of an additional argument that helps us to represent the data better. These types of arguments allow for better control over the plots that we create. To find what arguments each geom has, look for it in the Help section or put **?** in the `Console` in front of the function.

Geoms in `ggplot2` have a large selection of arguments to allow you to adjust your visualisations. This is what makes it one of the most preferred visualisation packages by the R community.


# Combination Plots

Imagine you want to see the difference in the number of people coming into and moving out of Scotland from both the RUK and Overseas throughout the years, but you do not want to know the net flow of people, their age or their sex. 

Q: What Variables would you use?

**Write your answer here**
A:


First, let's get the data we will need. We want our selected variables from above, and the overall data for the rest of the variables i.e. the 'all' categories for `Age`, and `Sex`. Remember we want to look at people moving in and out, so we do not want the net `Migration Type`. Save the output into the variable **traffic_scot2** so that it appears in your `Global Environment`.

```{r}
#first let's get the data we will need
# We want our chosen variables, and the overall data for the rest 
traffic_scot2 <- migration_scot %>%
  filter(Sex == 'All', 
         Age == 'All', 
         Migration_Type != 'Net') %>%
  select(DateCode, Value, Migration_Source, Migration_Type) 
```


Q: What type of data are these variables?

q1: DateCode             a1: 
q2: Value                a2: 
q3: Sex                  a3: 
q4: Migration Source     a4: 
q5: Migration Type       a5: 


So we need to use a different type of plot - one that can portray continuous data well enough.

Good way to portray data like this are point graphs (i.e. scatter plots) and line graphs.

We have several variables again. It will help if we separate them with different aesthetics.


### Scatter + Line

```{r}
ggplot(traffic_scot2, aes(DateCode, Value, shape=Migration_Source, color=Migration_Type)) + 
  geom_point()

```

This seems difficult to look at. Let's make it easier by adding a line that connects the data. We can do this by adding another **Layer**.


```{r}
ggplot(traffic_scot2, aes(DateCode, Value,
                          shape=Migration_Source)) + 
  geom_point() + 
  geom_line(aes(color=Migration_Type))

```


## Violin + Boxplots

We need a continuous variable and a categorical variable for both of them. In both of these the categorical variable usually goes on the x-axis and the continuous on the y axis.

Let's get back to the original data and plot the distribution of all females entering and leaving Scotland from overseas, from all ages. For this, we need to tidy the data into what we need and store it into the variable `female_migration`

```{r}
female_migration <- migration_scot %>%
  filter(Sex=='Female',
         Migration_Type!='Net',
         Migration_Source=='To-from Overseas',
         Age=='All') %>%
  select(Value, Migration_Type)

# make box/violin plots
ggplot(female_migration, aes(Migration_Type,Value)) + 
  geom_violin(aes(fill=Migration_Type)) +
  geom_boxplot(width=0.1) +
  geom_jitter()

```


## Practice Task 1


Using the data for this class follow the instructions. Remember, you can use `View()` in the console to look at your data, or click on the data variable in your `Global Environment`. This can help you to make your choices for the steps bellow and with the exact spelling. Start from the original dataset we loaded in the beginning `migration_scot`. Use pipes `%>%` for the first 5 steps and save the output into a variable `my_data`. Use that data to create three separate plots for steps 6, 7 and 8.

1. Select two different years. 
2. Select two different age groups. 
3. Select the total for both sex ( i.e. 'all').
4. Select migration In and Out of Scotland.
5. Pick whether you are going to look at `To-from Rest of UK` or `To-from Overseas`.
6. Plot the overall people, regardless of age and year, that have moved into and left Scotland using `geom_col()`. Save the plot in the variable q6 into your `Global Environment`.
7. Create a new plot using `geom_col()` that shows the differences between the two years you have selected, regardless of age. Save the plot in the variable q7 into your `Global Environment`.
8. Plot the differences between the two ages, regardless of year. 


```{r}
# do any wrangling you need to get the data you need
# Check the solution online for an Example answer
my_data <- migration_scot %>%

q6 <- ggplot(NULL) + geom_?
q6

q7 <- ggplot(NULL) + geom_?
q7 

q8 <-ggplot(NULL) + geom_?
q8


```

## Global and Local aesthetics
So far, we have been working with a Global definition of the aesthetics. This allows us to specify the axes and the groupings only once. 

Sometimes, we want to use multiple data and put multiple plots on top of each other. For these occasions we can specify the aesthetics at a local level - i.e. in the geoms.

If you remember from last week, we specified both the colour and the shape in our `ggplot()` when we were making the line graph. The `geom_line()` inherited the colour but not the shape argument. 

However, if we have two geoms that share the same aesthetics but we do not want to have the same colours, we can move them to each of the geoms instead of specifying them in `ggplot()`.

Let's redo the violins and box plots from last week, and put them together into two separate layers of the same graph. We want one boxplot-violin combo for each category within MigrationType. This time, we will give them some colour. We will colour the violins with the `fill` argument and make the box plots transparent using the `alpha` argument. We can also make the boxplots a little smaller using the `width` argument, so that they fit nicely over the violin plots. 

NOTE: We only need to use `aes()` if we are *mapping* something. What does this mean? Mapping simply refers to applying some process to a series of items. In this case, we are mapping the `fill` process to all of the unique values within MigrationType (which would be 'In' and 'Out' in this case). If you are simply specifying a single, specific value, you do not need to put this inside `aes()`, but rather just as an argument by itself. Let's test this out.

To get the data in the correct format, we need the distribution of *all females* entering *and* leaving Scotland *from overseas*, from *all ages*.


For the plot:
- We want one plot with 2 violin plots for the two categories in MigrationType (In and Out)
- Colour the violin plots based on whether they are showing In or  Out
- For each violin, we want a boxplot on top that shows the same data
- Adjust the 'width' argument of the boxplots so that they fit nicely in the violins
- Adjust the transparency of the boxes using the alpha argument


```{r}
#some data wrangling
boxes <- migration_scot %>%
  filter(Sex == 'Female',
         MigrationType != 'Net',
         MigrationSource == 'To-from Overseas',
         Age == 'All')
  

#make violins and box plots 
ggplot(boxes, aes(MigrationType, Value, fill=MigrationType)) + 
  geom_violin() +
  geom_boxplot(width=0.2, alpha=0.5, fill='White')
  
```

Because `alpha` (the transparency argument) and `fill` are arguments for both box plots and violins, if we had put them in the `ggplot()` layer, they would both be inherited by the two geoms.

Let's see how it would have looked like if we had specified the two arguments globally. 

```{r}
#make violins and box plots 
ggplot(boxes, aes(x = MigrationType, y = Value, fill = MigrationType, alpha = 0.5)) + 
  geom_violin() + 
  geom_boxplot()

```

Now the violins and the box plots are both coloured, which actually makes the transparency redundant. We could achieve the same effect as before by specifying `fill = 'white'` inside `geom_boxplot()`, which would override the global aesthetic. 



## Sub-Plots

Sometimes, we have way too many variables and they are all important. You should always avoid having busy plots. They get too confusing, and it is easy for people to misinterpret them.

We are going to look at two ways to create multiple sub-plots.

### Method 1: Patchwork

Create 2 different plot objects and combine them using the `patchwork` package.

#### Select the data before hand and create different plots

Simple Task: 
  1. From the data we just created, select only data relevant for movement to and from the rest of the UK and save it to the `Global Environment` as an object called `rest_of_uk` .

```{r}
# select is a bit misleading! maybe read it as 'retain only data relevant...'
rest_of_uk <- traffic_scot %>% 
  filter(MigrationSource == 'To-from Rest of UK')
  

```


 2. Create a column plot that shows the number of people for each sex using the data we just created using `geom_col()`. Separate the data by using different colours based on MigrationType. Don't forget to use `position = 'dodge'` so that your columns do not overlap. 


```{r}
rest_of_uk_plot <- ggplot(rest_of_uk, aes(x = Sex, y = Value, 
                       fill = MigrationType)) + 
  geom_col(position = 'dodge')
rest_of_uk_plot
```
 


 3. Now plot only the data for the movement to and from Overseas following the previous two steps. This time, make sure you save your wrangled data into an object called `overseas` and then use it to create your plot.

 
```{r}
overseas <- overseas<- traffic_scot %>% 
  filter(MigrationSource == 'To-from Overseas')

overseas_plot <- ggplot(overseas, aes(x = Sex, y = Value, 
                     fill = MigrationType)) + 
  geom_col(position = 'dodge')
overseas_plot
```
 
#### Combine with Patchwork

The `patchwork` package enables us to very simply combine plots using `+` and `/`. We can put them side by side:

```{r}
rest_of_uk_plot + overseas_plot
```

Or one above the other: 

```{r}
rest_of_uk_plot / overseas_plot
```

Or even a mix of the two:

```{r}
rest_of_uk_plot / (overseas_plot + overseas_plot)
```

This is very useful, but can become cumbersome when we have more than one variable. Imagine you have data for 10 different countries. Can you imagine doing these plots by hand? Moreover, when you have to arrange them on a page to show how things differ you will lose a lot of time. Instead we can use _facets_.
 
### Facets

Facets allow us to create separate sub-plots of different combinations of variables without manually separating the data. Moreover we can specify how we want to put the plots in a grid: do we want them side by side or one on top of each other, etc. 
 
We can use two main facet functions: 
 `facet_grid(rows, cols)` structures the rows and columns of graphs based on additional specified variables in `rows` and `cols`
 `facet_wrap(facets, nrows, ncols)` creates a 1-dimensional ribbon of plots based on NON-MISSING inputs to `facets`, and converts it to 2-dimensions based on `nrow` and `ncol`
 

### Wrap vs Grid

This can be a confusing distinction, as they often produce very comparable outputs, but let's try to see how they differ with a similar example. Let's look at the fuel economy data from the built in dataset `mpg`, and compare city vs highway:

```{r}
head(mpg)

grid_example <- ggplot(mpg, aes(cty, hwy)) #save the base plot structure
```


The key thing to note is that `facet_grid()` ALWAYS creates a grid of the specified dimensions, and then fills the grid with plots if possible. This means that you could have a grid with some blank plots if there are some missing values for certain combinations.

```{r}
grid_example + 
  geom_point() + 
  facet_grid(rows = vars(cyl), cols = vars(class))
```
We can see that certain combinations don't exist in the dataset (e.g. 5 cyl minivan), but the plot still shows up to preserve the grid format. Compare this with `facet_wrap()`, which just lines up all of the plots with data one after the other, then displays them in whatever format is specified (with a similar grid-like default).

```{r}
grid_example + 
  geom_point() + 
  facet_wrap(vars(cyl, class), ncol = 6)
```


Notice that we do not fill up the full dimensions of the grid, as we have specified 6 cols and only have 19 non-missing combinations of `cyl` and `class`. To better understand the logic, let's look at how this progresses if we use rows instead, starting with 1 row and expanding out:


```{r}
grid_example + 
  geom_point() + 
  facet_wrap(vars(cyl, class), nrow=1)
```

```{r}
grid_example + 
  geom_point() + 
  facet_wrap(vars(cyl, class), nrow=2)
```

```{r}
grid_example + 
  geom_point() + 
  facet_wrap(vars(cyl, class), nrow=3)
```

```{r}
grid_example + 
  geom_point() + 
  facet_wrap(vars(cyl, class), nrow=4)
```

We can see that `facet_wrap` starts with a single row of plots, and then proceeds to fill out new columns or rows based on the dimensions specified.

### Applying Facets to our Example

Let's replicate the two plots we just created but using `facet_grid()` and put them side-by-side in one. Now we have both plots next to each other, which makes it much easier to look at the differences in the migration patterns between the two Migration Sources.
 
```{r}
ggplot(traffic_scot, aes(x = Sex, y = Value, fill = MigrationType)) + 
  geom_col(position = 'dodge') +
  facet_grid(cols = vars(MigrationSource))

```

Let's look at these plots: 

Q1) Which sex tends to move to Scotland more, regardless of migration location? 
A1: F

Q2)  Which migration location do both males and females migrate to more when they leave Scotland? 
A2: RUK


#Themes: Making your plots pretty

The great thing about making your own plots in R is that you can make them look the way you want. Let's adjust the look of our graph.

## Ready made themes

ggplot2 comes with several ready made themes. Some of them are: 

 - theme_bw(): a white background with major axes and border
 - theme_minimal(): a white background with major axes and  more 
 Just use the cheatsheet for visualisations with ggplot for some more examples

Themes are just like any other layer for ggplot. You just add them at the end of your plot with `+`. Let's add one of these theme formats to our plot above. 

```{r}
ggplot(boxes, aes(MigrationType, Value)) + 
  geom_violin(aes(fill = MigrationType)) + 
  geom_boxplot(alpha = 0.5, width = 0.2) +
  theme_light()
```


## Adding axes titles and headings

Next, let's add some titles and better labels with `labs()`. You need to specify which labels you want to change - x, y or the title, and type in the labels you want. Write `?labs()` in your console to see some examples.

```{r}
ggplot(boxes, aes(x = MigrationType, y = Value)) + 
  geom_violin(aes(fill = MigrationType)) + 
  geom_boxplot(alpha = 0.5, width = 0.2) +
  theme_bw()+
  labs(x = 'Migration Direction',
       y = 'Number of People',
       title = 'Female Migration between Scotland and Overseas') # your code here

```

Sometimes, we need the titles to be a specific size. We can change all of that in an additional `theme()` layer. We specify the argument we want to change and what about them we need to change. In our case, I want to change the *size of the x-axis title to 12pt*, and *the graph title to 16pt*. Let's also pretend we *do not want the y-axis title*. We can also specify this in the `theme()` layer. 

We do this by specifying `element_text()` and `element_blank()` for the correct theme arguments (`axis.title.x`, `title`, `axis.title.x`). Simply put, for that element of the plot, what do we want to change - text and to make it blank.

```{r}
ggplot(boxes, aes(x = MigrationType, y = Value)) + 
  geom_violin(aes(fill = MigrationType)) + 
  geom_boxplot(alpha = 0.5, width = 0.2) +
  theme_bw() + 
  labs(title = 'Female Migration between Scotland and Overseas', 
       x = 'Migration direction', 
       y = 'Number of people') + 
  theme(
    axis.title.x = element_text(size=12), # your code here
    title = element_text(size=16), # your code here
    axis.title.y =  element_blank()# your code here
  )

```


## Changing colours

We can rely on the colour schemes automatically set by `ggplot2`, or we can specify colours we want. In our case, we are changing the colours of the `fill` aesthetic, so we could use `scale_fill_manual()`. Colours are adjusted using the `values` argument, which takes a vector of color names or hex codes.  The `scale_fill_manual()` function is implemented as an additional layer and it overwrites the default colour scheme. Because of this, it has to be specified *after* the geom that uses the `fill` argument.

```{r}
ggplot(boxes, aes(x = MigrationType, y = Value)) + 
  geom_violin(aes(fill = MigrationType)) + 
  geom_boxplot(alpha = 0.5, width = 0.2) +
  theme_bw() + 
  labs(title = 'Female Migration between Scotland and Overseas', 
       x = 'Migration direction', 
       y = 'Number of people') + 
  theme(
    axis.title.x = element_text(size = 12),
    title = element_text(size = 16),
    axis.title.y = element_blank()
  ) +
 scale_fill_manual(values = c('gold','dodgerblue')) #your code here

```

## Legends

We can see that for any additional aesthetic we have put, ggplot has created a legend on the side.

Now let's change what the labels actually say. We do this in the same place we specified the colours `scale_fill_manual` because, again, we are overwriting the automatic `fill` labels. 

Let's also: 

1) hide the legend associated with the box plots (i.e. `alpha`) using `legend.position = FALSE`; 

2) move the legend to the bottom of the graph using the `legend.position` argument and specifying 'bottom';  

3) change the size of the legend titles (using `legend.title`). 


We do steps 2 and 3 in the `theme()` layer.

Because we have more than one legend, everything we do will be applied to both legends.This is a good idea for consistency. But when we want to hide one of the legends, we have to specify that in the geom where the legend is created using the argument `show.legend`.

```{r}
ggplot(boxes, aes(x = MigrationType, y = Value)) + 
  geom_violin(aes(fill = MigrationType)) + 
  geom_boxplot(alpha = 0.5, width = 0.2) + # add code to this line
  theme_bw() +
  labs(title = 'Female migration to and from\nOverseas into Scotland', x = 'Migration direction', y = 'Number of people') +
  theme(
    axis.title.x = element_text(size = 12),
    title = element_text(size = 16),
    axis.title.y = element_blank(),
    legend.position = 'right', # change code here 
    legend.title = element_text() # add change code here 
  ) +
  scale_fill_manual(values = c('gold','dodgerblue'),
                    name = 'Migration Direction',
                    labels = c('Coming to Scotland','Leaving Scotland'))

```


## Changing Axis Ticks

Now, we also need to change what is stated on our x-axis to match the legend. We use `scale_x_discrete()` to give new names inside the graph, so that we do not have to change the names within our dataset. We specify the labels of the new groups by giving a *vector* with c('old name' = 'new name') for each item we wish to change. We use `scale_x_discrete()`, because we have discrete groups - a categorical variable. 

```{r}
ggplot(boxes, aes(x = MigrationType, y = Value)) + 
  geom_violin(aes(fill = MigrationType)) + 
  geom_boxplot(alpha = 0.5, width = 0.2, show.legend = FALSE) +
  theme_bw() +
  labs(title = 'Female migration to and from\nOverseas into Scotland', x = 'Migration direction', y = 'Number of people') +
  theme(
    axis.title.x = element_text(size = 12),
    title = element_text(size = 20),
    axis.title.y = element_blank(),
    legend.position = 'bottom',
    legend.title = element_text(size = 10)
  ) +
  scale_fill_manual(values = c('gold','dodgerblue'),
                    name = 'Migration Direction',
                    labels = c('Coming to Scotland','Leaving Scotland')) +
  scale_x_discrete(labels = c(
    'In' = 'To Scotland',
    'Out' = 'From Scotland')) # your code here
  
```


Let's also say that we are plotting data which is continuous but we want to show each of the values. For example, we might want to show each year from the column `DateCode`. In this case, `ggplot` will skip some of the years to make the graph look neater. There are several ways we can change that. We can either make the variable into a factor or a character using `as.character` so that `ggplot` will be forced take it as a discrete variable, or we can change the ticks of the x axis by giving them `limits` and `breaks` in the layer that overwrites the original x values - `scale_x_continuous()`.


```{r}
ggplot(traffic_scot2, aes(DateCode, Value,
                          shape = MigrationSource, 
                          colour = MigrationType)) + 
  geom_point() + 
  geom_line() +
  scale_x_continuous(limits = c(2002,2018), breaks = 2002:2018)
```


## Saving plots

Finally, we have made these lovely plots but we want to be able to put them in reports, papers, on a poster (print it in a large canvas for friends and family to admire). To save your plot to your working directory, you can use the function `ggsave`. First, you have to save your plot into an object in the `Global Environment`. 

```{r}
# Save your plot to the Global Environment.
plot1 <- ggplot(traffic_scot2, aes(DateCode, Value,
                          shape = MigrationSource, 
                          colour = MigrationType)) + 
  geom_point() + 
  geom_line() +
  theme_bw() 

plot1
# Then save it to your working directory folder
ggsave('myplot.png', plot1, width = 9 , height = 6)

```


## Bonus Challenge! 

Similar to last week's challenge, your goal is to represent the same data in the most pretty and ugly way possible. This time, you have more tools at your disposal to customise different features of your plots. For this challenge, you must visualise how emigration numbers vary across age for males and females separately. You can choose any types or combinations of geoms, as long as it demonstrates this information. Remember, you can always refer to the cheatsheet, the help function (e.g. ?geom_boxplot), or the internet in general. Some intitial data wrangling has been done for you, but you can adjust this if it suits your plot idea. Have fun! 


```{r}
# data wrangling 
plot_data <- migration_scot %>%
  filter(
    Age != 'All',
    Sex != 'All',
    MigrationType == 'Out'
  ) %>%
  select(Value, Age, Sex) %>%
  group_by(Age, Sex) %>%
  summarise(Value = sum(Value)) 
# remove characters and convert to numeric
plot_data$Age <- as.numeric(gsub(' year','', gsub(' years','',plot_data$Age)))

```

```{r}
# pretty plot 1
pretty_plot1 <- ggplot(plot_data, aes(Age, Value, fill = Sex, alpha = Sex)) +
  geom_col(position = 'identity', width = 1, color='black') +
  scale_fill_manual(values=c('midnightblue','azure')) +
  theme_minimal() +
  labs(y = 'Number of Emigrations') + 
  scale_alpha_discrete(range=c(1,0.7))

# add column with negative values for Female data
plot_data$Value2 = plot_data$Value
plot_data$Value2[plot_data$Sex=='Female'] = plot_data$Value2[plot_data$Sex=='Female']*-1  

# pretty plot 2
pretty_plot2 <- ggplot(plot_data, aes(Age, Value2, fill = Sex)) +
  geom_col(position = "identity", width = 1) +
  scale_fill_brewer(palette = 'Accent') +
  theme_minimal() +
  coord_flip() +
  scale_y_continuous(
    breaks = seq(-30000,20000,10000),
    labels = c(30000,20000,10000,0,10000,20000)
  ) +
  labs(y = 'Number of Emigrations')

gridExtra::grid.arrange(pretty_plot1,pretty_plot2)

```

```{r}
#ugly_plot
ugly_plot <- ggplot(plot_data, aes(Age, Value, color = Sex)) +
  geom_point(size = 15, fill = 'white', aes(shape = Sex)) +
  geom_line() +
  scale_color_manual(values=c('salmon2','olivedrab4')) +
  theme_dark() +
  labs(y = 'Number of Emigrations') +
  theme(
    axis.title.x = element_text(size = 25),
    axis.line = element_line(color='darkorchid2',size=3),
    axis.title.y = element_text(size=16),
    legend.position = 'bottom',
    legend.title = element_text(size = 35)
  )

ugly_plot

```


